The Impact of Alternation

### 2017-05-15 draft.tex
ok

The execution of Algorithm \ref{alg:impact} consists of three
phases\footnote{These phases correspond to the \textsc{Close},
  \textsc{Refine} and \textsc{Expand} procedures from the original
  description of \impact~\cite{mcmillan06}.}: \emph{close},
\emph{refine} and \emph{expand}. Let $n$ be a node removed from the
worklist at line \ref{ln:impact-dequeue}. If $\Accept{\A}(\lambda(n))$
is satisfiable, the counterexample $\lambda(n)$ is feasible, in which
case a model of $\Accept{\A}(\lambda(n))$ is obtained and a word $w
\in L(\A)$ is returned. Otherwise, $\lambda(n)$ is a spurious
counterexample and the procedure enters the refinement phase (lines
\ref{ln:refine-begin}-\ref{ln:refine-end}). The interpolant for
$\Theta(\lambda(n))$ (cf. formula \ref{eq:interpolation-problem}) is
used to strenghten the labels of all the ancestors of $n$, by
conjoining the formulae of the interpolant to the existing labels.

@@

\begin{figure}[htb]
\vspace*{-\baselineskip}
\begin{tabular}{c}
\includegraphics[scale=0.65]{figures/IM1.pdf} \\[-2mm]
\tiny{(a)} \\
\includegraphics[scale=0.65]{figures/IM2.pdf} \\[-2mm]
\tiny{(b)} \\
\includegraphics[scale=0.65]{figures/IM3.pdf} \\[-2mm]
\tiny{(c)}
\end{tabular}
\vspace*{-0.5\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
  by Algorithm \ref{alg:impact} (1/2)}
\label{fig:im1}
\end{figure}

@@

The correctness of Algorithm \ref{alg:impact} is proved in the
following:

@@

We have implemented both Algorithm \ref{alg:predabs} and
\ref{alg:impact} in a prototype
tool\footnote{\url{https://github.com/cathiec/AltImpact}} that uses
the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} to
check satisfiability queries and generate interpolants in the theory
of linear integer arithmetic variables (LIA). We compared both
algorithms with a previous implementation of trace inclusion
procedure, called
\textsc{Includer}\footnote{\url{http://www.fit.vutbr.cz/research/groups/verifit/tools/includer/}},
that uses on-the-fly determinisation and lazy predicate abstraction
with interpolant-based refinement \cite{Tacas16}.

The results of the experiments are given in Table
\ref{tab:experiments}. We applied the tool on several array logic
entailments, which occur as verification conditions for imperative
programs with arrays \cite{cav09} (Array shift, Array rotation 1+2,
Array split) available online \cite{ntslib}. Then, we applied it on
proving safety properties of hardware circuits (HW Counter 1+2)
\cite{smrcka}. Finally, we considered two timed communication
protocols, consisting of systems that are asynchronous compositions of
timed automata, whom correctness specification are given by timed
automata monitors: two versions (correct and faulty) of the timed
Alternating Bit Protocol \cite{abp} and a~controller of a railroad
crossing \cite{henzinger:RealTimeSystems}. All results were obtained
on an Intel(R) Core(TM) i7-4650U CPU @ 1.70GHz with 8GB of
RAM. Execution times are given in seconds.

\begin{table}[htb]
\begin{center}
{\fontsize{8}{9}\selectfont
\begin{tabular}{||l|c|c|c|c|c||}
\hline
Example & $\len{\A}$ & $L(\A)=\emptyset$ ? & Algorithm \ref{alg:predabs} & Algorithm \ref{alg:impact} & \textsc{Includer} \\
\hline 
\hline
\end{tabular}
}
\caption{}\label{tab:experiments}
\end{center}
\end{table}

>>>>

The execution of Algorithm \ref{alg:impact} consists of three
phases\footnote{Corresponding to the \textsc{Close}, \textsc{Refine}
  and \textsc{Expand} in \cite{mcmillan06}.}: \emph{close},
\emph{refine} and \emph{expand}. Let $n$ be a node removed from the
worklist at line \ref{ln:impact-dequeue}. If $\Accept{\A}(\lambda(n))$
is satisfiable, the counterexample $\lambda(n)$ is feasible, in which
case a model of $\Accept{\A}(\lambda(n))$ is obtained and a word $w
\in L(\A)$ is returned. Otherwise, $\lambda(n)$ is a spurious
counterexample and the procedure enters the refinement phase (lines
\ref{ln:refine-begin}-\ref{ln:refine-end}). The interpolant for
$\Theta(\lambda(n))$ (cf. formula \ref{eq:interpolation-problem}) is
used to strenghten the labels of all the ancestors of $n$, by
conjoining the formulae of the interpolant to the existing labels.

@@

\begin{figure}[htb]
\vspace*{-\baselineskip}
\begin{tabular}{c}
\includegraphics[scale=0.65]{figures/IM1.pdf} \\
\tiny{(a)} \\
\includegraphics[scale=0.65]{figures/IM2.pdf} \\[-2mm]
\tiny{(b)} \\
\includegraphics[scale=0.65]{figures/IM3.pdf} \\[-2mm]
\tiny{(c)}
\end{tabular}
\vspace*{-0.5\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
  by Algorithm \ref{alg:impact} (1/2)}
\label{fig:im1}
\vspace*{-\baselineskip}
\end{figure}

@@

@@

We have implemented both Algorithm \ref{alg:predabs} and
\ref{alg:impact} in a prototype tool\footnote{The implementation is
  available at \url{https://github.com/cathiec/AltImpact}} that uses
the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} for
the satisfiability queries and interpolant generation, in the theory
of linear integer arithmetic (LIA) combined with booleans. We compared
both algorithms with a previous implementation of a trace inclusion
procedure, called
\textsc{Includer}\footnote{\url{http://www.fit.vutbr.cz/research/groups/verifit/tools/includer/}},
that uses on-the-fly determinisation and lazy predicate abstraction
with interpolant-based refinement \cite{Tacas16} in the LIA theory,
without booleans.

The results of the experiments are given in Table
\ref{tab:experiments}. We applied the tool first to several array
logic entailments, which occur as verification conditions for
imperative programs with arrays \cite{cav09} (array\_shift,
array\_simple, array\_rotation 1+2) available online
\cite{ntslib}. Next, we applied it on proving safety properties of
hardware circuits (hw1+2) \cite{smrcka}. Finally, we considered two
timed communication protocols, consisting of systems that are
asynchronous compositions of timed automata, whom correctness
specifications are given by timed automata monitors: a timed version of
the Alternating Bit Protocol (abp) \cite{abp} and a~controller of a
railroad crossing (train) \cite{henzinger:RealTimeSystems}. All
results were obtained on an Intel(R) Core(TM) i7-4650U CPU @ 1.70GHz
with 8GB of RAM. The automata sizes are in bytes and the execution
times are in seconds.

\begin{table}[htb]
\begin{center}
{\fontsize{8}{9}\selectfont
\begin{tabular}{||l|c|c|c|c|c||}
\hline
Example & $\len{\A}$ & $L(\A)=\emptyset$ ? & Algorithm \ref{alg:predabs} & Algorithm \ref{alg:impact} & \textsc{Includer} \\
\hline
simple1 & 309 & no & 0.641 & 0.061 & 0.021\\
\hline
simple2 & 504 & yes & 0.667 & 0.179 & 0.031\\
\hline
simple3 & 214 & yes & 0.825 & 0.150 & 0.051\\
\hline
array\_shift & 874 & yes & 2.138 & 0.209 & 0.060\\
\hline
array\_simple & 3440 & yes & timeout & 16.435 & 5.496\\
\hline
array\_rotation1 & 1834 & yes & 6.827 & 0.865 & 0.114\\
\hline
array\_rotation2 & 15182 & yes & timeout & timeout & 23.056\\
\hline
abp & 6909 & no & 8.292 & 1.410 & 1.715\\
\hline
train & 1823 & yes & 16.989 & 2.738 & 0.308\\
\hline
hw1 & 322 & yes & 1.566 & 0.298 & 0.159\\
\hline
hw2 & 674 & yes & 22.914 & 0.552 & 0.419\\
\hline
\end{tabular}
}
\caption{}\label{tab:experiments}
\end{center}
\end{table}

As in the non-deterministic case \cite{mcmillan06}, \impact~
outperforms lazy predicate abstraction for checking emptiness by at
least one order of magnitude. However, both our implementations are
slower than \textsc{Includer}, on average (except for the abp
example). The reason for this is currently under investigation, one
possible bottleneck being the hardness of the combined (LIA+booleans)
interpolation problems, as opposed to converting the entire formula
into DNF, eliminating the boolean variables and using interpolation in
the pure LIA theory.
